Position: Stop Making Unscientific AGI Performance Claims

### 2024-05-28 body.tex
proof-read one more time

At the same time, in the way in which academic research is disseminated in the AI and machine learning fields today, major shifts are happening. Where these research fields have actively adopted early sharing of preprints and code, the volume of publishable work has exploded to a degree that one cannot reasonably keep up with broad state-of-the-art, and social media influencers start playing a role in article discovery and citeability~\citep{weissburg2024tweets}. Furthermore, because of major commercial stakes with regard to AI and machine learning technology, and e.g.\ following the enthusiastic societal uptake of products employing large language models (LLMs), such as ChatGPT, the pressure to beat competitors as fast as possible is only increasing, and strong eagerness can be observed in many domains to `do something with AI' in order to innovate and remain current.

@@

Two other recent works are related to this work and align well with the position we present here.~\citet{schaeffer2024emergent} demonstrate that the apparent emergent abilities of large language models may be driven by a choice of evaluation metrics, rather than some fundamental property that is intrinsic to this family of models. Their work highlights the need for rigorous testing and benchmarking of LLMs, which we also point to in this work albeit in a slightly different methodological context.~\citet{kloft2024ai} provide experimental evidence demonstrating that people have heightened expectations and a biased, positive view of AI. The authors run a user study of human-AI interaction, in which participants performed better at a given task when they thought they were aided by a positively described AI. This Placebo effect was found to be robust to negative descriptions of AI. Positive bias towards AI may exacerbate other factors that drive people to make unscientific claims about the current state of AI, which we discuss in Section~\ref*{social}.

@@

This first experiment---and indeed the practice of probing LLMs that have seen vast amounts of data---can be seen as a form of inverse problem and common caveats such as non-uniqueness and instability apply~\citep{altmeier2023regularization}. Regularization can help mitigate these caveats~\citep{om2001ridge}, but we confess that we did not carefully consider the parameter choice for $\lambda $, nor has this been carefully studied in the related literature to the best of our knowledge.

@@

So far, we considered simple linear data transformations. One might argue that this does not really involve latent embeddings in the way they are typically thought of in the context of deep learning. In Appendix~\ref{appendix:autoencoder}, we present an additional experiment in which we more explicitly seek neural network-based representations that will be useful for downstream tasks. Here, we continue with an example in which LLMs may be used for Economic Sentiment Prediction.

Closely following the approach in \citet{gurnee2023languagev2}, we apply it to the novel \emph{Trillion Dollar Words}~\citep{shah2023trillion} financial dataset, containing a curated selection of sentences formulated by central bankers of the US Federal Reserve, and communicated to the public in speeches, meeting minutes and press conferences. ~\citep{shah2023trillion} use this dataset to train a set of LLMs and rule-based models to classify sentences as either `dovish', `hawkish' or `neutral'. In the context of central banking, `hawkishness' is typically associated with tight monetary policy: in other words, a `hawkish' stance on policy favors high interest rates to limit the supply of money and thereby control inflation. The authors first manually annotate a sub-sample of the available data and then fine-tune various models for the classification task. Their model of choice, \emph{FOMC-RoBERTa} (a fine-tuned version of RoBERTa \citep{liu2019roberta}), achieves an \(F_1\) score of around \(>0.7\) on the test data. To illustrate the potential usefulness of the learned classifier, they use predicted labels for the entire dataset to compute an ad-hoc, count-based measure of `hawkishness'. This measure is shown to correlate with key economic indicators in the expected direction: when inflationary pressures rise, the measured level of `hawkishness' increases, as central bankers react by raising interest rates to bring inflation back to target.

@@

Research on anthropomorphism has repeatedly shown the human tendency to attribute human-like characteristics to non-human agents and/or objects. These might include the weather and other natural forces, pets and other animals, gadgets and other pieces of technology~\citep{epley2007seeing}. Formally studied as early as 1944,~\citep{heider1944experimental} observed that humans can correctly interpret a narrative whose characters are abstract 2D shapes, but also that humans interpreted random movements of these shapes as having a human-like narrative. 
Relevant to AI and the degree to which it resembles AGI, anthropomorphizing may occur independently of whether such judgments are accurate, and as a matter of degree: at the weaker end, one may employ anthropomorphism as a metaphorical way of thinking or explaining, and at the stronger end one may attribute human emotions, cognition, and intelligence to AI systems. As \citet{epley2007seeing} note, literature has shown that even weak metaphorical anthropomorphism may affect how humans behave towards non-human agents.

Modern anthropomorphism theory suggests there are three key components, one of which is a cognitive feature, and two of which are motivations. The first involves the easy availability of our experiences as heuristics that can be used to explain external phenomena: ``...knowledge about humans in general, or self-knowledge more specifically, functions as the known and often readily accessible base for induction about the properties of unknown agents" (p.866) \citep{epley2007seeing}. Thus, our experience as humans is an always-readily-available template to interpret the world, including non-human agent behaviors.

@@

The third component is the motivation as a human to be competent (effectance motivation). This is most relevant to this discussion, as it describes the need to effectively interact with our environments, including the technologies of the day \citep{epley2007seeing}. When confronted with an opaque technology, a person may interpret its behaviors using the most readily available template at hand, namely their personal human experience, in order to facilitate learning \citep{epley2007seeing, waytz2010social}. Perceiving human characteristics, motivations, emotions, and cognitive processes from one's own experiences in a technology e.g. an AI chatbot, allows for a ready template of comparison at the very least, and possibly an increase in ability to make sense of, and even predict, the agent's behaviors. This may include being placed in a position to master a certain technology, whether by incentives to learn, or fear of poor outcomes should one not manage to learn.

These pressures extend to AI experts, as well as laypersons. In both scholarly and commercial fields, AI experts face considerable pressure to demonstrate competence in their work. Citation metrics and scholarly publications remain the primary metric for tenure and promotion, e.g. \citep{alperin2019significant}, and the number of publications in the AI field has boomed as evidenced by overall (preprint and peer-reviewed) scholarly publications\footnote{\url{https://ourworldindata.org/grapher/annual-scholarly-publications-on-artificial-intelligence?time=2010..2021} }~\citep{Maslej2023-pi}.

@@

Hypotheses in present-day AI research are often implicit. Generally, these hypotheses are framed simply as a system being more accurate or efficient, compared to other systems. Where other fields, such as medicine or quantitative social sciences, would further articulate expectations in e.g.\ assigning specific conditions and considering effect sizes assigned to each competing hypothesis, in computer science and AI this is typically not done. This also may have to do with much of the published work being more of an engineering achievement, rather than a true hypothesis test seeking to explain and understand the world. 
However, in discussions on emerging qualities like AGI, this engineering positioning gets muddier, and more formal hypothesis testing would be justifiable: either one interprets outputs as in support of hints towards AGI (the competing hypothesis), or as merely the result of an algorithm integrating qualities from the data it was trained on (the null hypothesis).

@@

Moreover, we believe that structural and cultural changes are in order to reduce current incentives to chase Big Statement Outcomes in AI research and industry. Our broadest and perhaps most ambitious goal is for our research community to move away from authorship and instead embrace contributorship. This argument has been raised long before both in other research communities~\citep{smith1997authorship} and more recently within our own~\citep{liem2023treat}. Specifically,~\citet{liem2023treat} argue that societally impactful scientific insights should be treated as open-source software artifacts. The open-source community sets a positive example of how scientific artifacts should be published in many different ways. Not only does it adequately reward small contributions but it also naturally considers negative results (bugs) as part of the scientific process. Similarly, code reviews are considered so integral to the process that they typically end up as accredited contributions to projects. Open review platforms like OpenReview are a step in the right direction but they still fall short of what is technically feasible. Finally, software testing is, of course, not only essential but often obligatory before contributions are accepted and merged. As we have pointed out repeatedly in this work, any claims about AGI demand proper strong hypothesis tests. It is important to remember that AGI remains the alternative hypothesis and that the burden of proof therefore lies with those making claims.

>>>>

At the same time, in the way in which academic research is disseminated in the AI and machine learning fields today, major shifts are happening. Where these research fields have actively adopted early sharing of preprints and code, the volume of publishable work has exploded to a degree that one cannot reasonably keep up with broad state-of-the-art, and social media influencers start playing a role in article discovery and citeability~\citep{weissburg2024tweets}. Furthermore, because of major commercial stakes with regard to AI and machine learning technology, and e.g.\ following the enthusiastic societal uptake of products employing LLMs, such as ChatGPT, the pressure to beat competitors as fast as possible is only increasing, and strong eagerness can be observed in many domains to `do something with AI' in order to innovate and remain current.

@@

Two other recent works are related to this work and align well with the position we present here.~\citet{schaeffer2024emergent} demonstrate that the apparent emergent abilities of large language models may be driven by a choice of evaluation metrics, rather than some fundamental property that is intrinsic to this family of models. Their work highlights the need for rigorous testing and benchmarking of LLMs, which we also point to in this work, albeit in a slightly different methodological context.~\citet{kloft2024ai} provide experimental evidence demonstrating that people have heightened expectations and a biased, positive view of AI. The authors run a user study of human-AI interaction, in which participants performed better at a given task when they (wrongly) thought they were aided by a positively described AI. This Placebo effect was found to be robust to negative descriptions of AI. Positive bias towards AI may exacerbate other factors that drive people to make unscientific claims about the current state of AI, which we discuss in Section~\ref*{social}.

@@

This first experiment---and indeed the practice of probing LLMs that have seen vast amounts of data---can be seen as a form of inverse problem and common caveats such as non-uniqueness and instability apply~\citep{haltmeier2023regularization}. Regularization can help mitigate these caveats~\citep{om2001ridge}, but we confess that we did not carefully consider the parameter choice for $\lambda $, nor has this been carefully studied in the related literature to the best of our knowledge.

@@

So far, we considered simple linear data transformations. One might argue that this does not really involve latent embeddings in the way they are typically thought of in the context of deep learning. In Appendix~\ref{appendix:autoencoder}, we present an additional experiment in which we more explicitly seek neural network-based representations that will be useful for downstream tasks. Here, we continue with an example in which LLMs may be used for economic sentiment prediction.

Closely following the approach in \citet{gurnee2023languagev2}, we apply it to the novel \emph{Trillion Dollar Words}~\citep{shah2023trillion} financial dataset, containing a curated selection of sentences formulated and communicated to the public by the Fed through speeches, meeting minutes and press conferences.~\citep{shah2023trillion} use this dataset to train a set of LLMs and rule-based models to classify sentences as either `dovish', `hawkish' or `neutral'. In the context of central banking, `hawkishness' is typically associated with tight monetary policy: in other words, a `hawkish' stance on policy favors high interest rates to limit the supply of money and thereby control inflation. The authors first manually annotate a sub-sample of the available data and then fine-tune various models for the classification task. Their model of choice, \emph{FOMC-RoBERTa} (a fine-tuned version of RoBERTa \citep{liu2019roberta}), achieves an \(F_1\) score of around \(>0.7\) on the test data. To illustrate the potential usefulness of the learned classifier, they use predicted labels for the entire dataset to compute an ad-hoc, count-based measure of `hawkishness'. This measure is shown to correlate with key economic indicators in the expected direction: when inflationary pressures rise, the measured level of `hawkishness' increases, as central bankers react by raising interest rates to bring inflation back to target.

@@

Research on anthropomorphism has repeatedly shown the human tendency to attribute human-like characteristics to non-human agents and/or objects. These might include the weather and other natural forces, pets and other animals, gadgets and other pieces of technology~\citep{epley2007seeing}. Formally studied as early as 1944,~\citet{heider1944experimental} observed that humans can correctly interpret a narrative whose characters are abstract 2D shapes, but also that humans interpreted random movements of these shapes as having a human-like narrative. 
Relevant to AI and the degree to which it resembles AGI, anthropomorphizing may occur independently of whether such judgments are accurate, and as a matter of degree: at the weaker end, one may employ anthropomorphism as a metaphorical way of thinking or explaining, and at the stronger end one may attribute human emotions, cognition, and intelligence to AI systems. As \citet{epley2007seeing} note, literature has shown that even weak metaphorical anthropomorphism may affect how humans behave towards non-human agents.

Modern anthropomorphism theory suggests there are three key components, one of which is a cognitive feature, and two of which are motivations. The first involves the easy availability of our experiences as heuristics that can be used to explain external phenomena: ``...knowledge about humans in general, or self-knowledge more specifically, functions as the known and often readily accessible base for induction about the properties of unknown agents" (p.866 in \citet{epley2007seeing}). Thus, our experience as humans is an always-readily-available template to interpret the world, including non-human agent behaviors.

@@

The third component is the motivation as a human to be competent (effectance motivation). This is most relevant to this discussion, as it describes the need to effectively interact with our environments, including the technologies of the day \citep{epley2007seeing}. When confronted with an opaque technology, a person may interpret its behaviors using the most readily available template at hand, namely their personal human experience, in order to facilitate learning \citep{epley2007seeing, waytz2010social}. Perceiving human characteristics, motivations, emotions, and cognitive processes from one's own experiences in a technology such as an AI chatbot, allows for a ready template of comparison at the very least, and possibly an increase in ability to make sense of, and even predict, the agent's behaviors. This may include being placed in a position to master a certain technology, whether by incentives to learn, or fear of poor outcomes should one not manage to learn.

These pressures extend to AI experts, as well as laypersons. In both scholarly and commercial fields, AI experts face considerable pressure to demonstrate competence in their work. Citation metrics and scholarly publications remain the primary metric for tenure and promotion~\citep{alperin2019significant}, and the number of publications in the AI field has boomed as evidenced by overall (preprint and peer-reviewed) scholarly publications\footnote{\url{https://ourworldindata.org/grapher/annual-scholarly-publications-on-artificial-intelligence?time=2010..2021} }~\citep{Maslej2023-pi}.

@@

Hypotheses in present-day AI research are often implicit. Generally, these hypotheses are framed simply as a system being more accurate or efficient, compared to other systems. Where other fields, such as medicine or quantitative social sciences, would further articulate expectations in e.g.\ assigning specific conditions and considering effect sizes assigned to each competing hypothesis, in Computer Science and AI this is typically not done. This also may have to do with much of the published work being more of an engineering achievement, rather than a true hypothesis test seeking to explain and understand the world. 
However, in discussions on emerging qualities like AGI, this engineering positioning gets muddier, and more formal hypothesis testing would be justifiable: either one interprets outputs as in support of hints towards AGI (the alternative hypothesis), or as merely the result of an algorithm integrating qualities from the data it was trained on (the null hypothesis).

@@

Moreover, we believe that structural and cultural changes are in order to reduce current incentives to chase Big Statement Outcomes in AI research and industry. Our broadest and perhaps most ambitious goal is for our research community to move away from authorship and instead embrace contributorship. This argument has been raised long before in other research communities~\citep{smith1997authorship} and more recently within our own~\citep{liem2023treat}. Specifically,~\citet{liem2023treat} argue that societally impactful scientific insights should be treated as open-source software artifacts. The open-source community sets a positive example of how scientific artifacts should be published in many different ways. Not only does it adequately reward small contributions but it also naturally considers negative results (bugs) as part of the scientific process. Similarly, code reviews are considered so integral to the process that they typically end up as accredited contributions to projects. Open review platforms like OpenReview are a step in the right direction but they still fall short of what we know is technically feasible. Finally, software testing is, of course, not only essential but often obligatory before contributions are accepted and merged. As we have pointed out repeatedly in this work, any claims about AGI demand proper strong hypothesis tests. It is important to remember that AGI remains the alternative hypothesis and that the burden of proof therefore lies with those making strong claims.
